---
title: "glinski_metabolites"  
output: html_document
---
tested hypotheses
============
- observed vivo versus in vitro degradation rates are the same, or observed vivo rates are lower than in vitro rates (due to other degradation processes, uptake kinetics?)
- first order kinetics are michaelis-menton, second order?

things to do
============
- load actual microsome data into csv
- negative and zero concentration values in exposure experiment data
- michaelis menton first order comparison for both experiments

general configuration
=====================
Install and load supporting libraries.

```{r eval=TRUE, echo=FALSE}
print(Sys.info()[4])

#original data sources are available on dropbox
#/amphib_glinski_micosomes

#install.packages('rmarkdown')
#install.packages(dplyr)
#install.packages(knitr)
#install.packages(zoo)
#install.packages(xtable)
#install.packages(drc) # for fitting Michaelis Menten model
#install.packages(ggplot2) # for drawing
#library(grofit)

R.Version()$version.string
library(rmarkdown, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(knitr, quietly = TRUE, warn.conflicts = FALSE)
library(zoo, quietly = TRUE, warn.conflicts = FALSE)
library(xtable, quietly = TRUE, warn.conflicts = FALSE)
library(drc, quietly = TRUE, warn.conflicts = FALSE) # for fitting Michaelis Menten model
library(ggplot2, quietly = TRUE, warn.conflicts = FALSE) # for drawing
#library(grofit)
print("list of loaded packages: ")
print((.packages()))
```

import experimental exposure data
=================================
Load csv file with experimental data.

```{r eval=TRUE, echo=FALSE}
if(Sys.info()[4]=="DONNA-PC"){
  micro.root <- path.expand("C:/Users/Donna/Documents/Github/glinski_metabolites/")
}

if(Sys.info()[4]=="stp-air-3.local" || Sys.info()[4]=="stp-air.local" || Sys.info()[4]=="stp-air"){
  micro.root <- path.expand("~/git/glinski_metabolites/")
}

if(Sys.info()[4]=="DZ2626UTPURUCKE"){
  micro.root <- "k:/git/glinski_metabolites/"
}
print(paste("Root directory location: ", micro.root, sep=""))

micro.csv.in <- paste(micro.root, "csv_in/", sep="")
micro.csv.out <- paste(micro.root, "csv_out/", sep="")
micro.graphics <- paste(micro.root, "graphics/", sep="")
micro.tables <- paste(micro.root, "tables/", sep="")

#maybe ok even if remote drive returns false
print(paste("check to see if R can access files OK: ", file.exists(micro.csv.in), sep = ""))

#import data
micro <- read.table(paste(micro.csv.in,"exposure_experiment.csv",sep=""), header = TRUE, sep = ",")
microsome <- read.table(paste(micro.csv.in,"microsome_experiment.csv",sep=""), header = TRUE, sep = ",")
```

data structure
==============
Check out structure of imported data set. 

```{r eval=TRUE, echo=TRUE}
str(micro)
```

Set time and tank fields as factor for later statistical inference.
```{r eval=TRUE, echo=FALSE}
micro$time <- factor(micro$time)
micro$replicate <- factor(micro$replicate)
parents <- unique(micro$parent)
analytes <- unique(micro$analyte)

micro
parents
analytes

micro.amphib <- micro[which(micro$matrix=="amphib"),]
micro.soil <- micro[which(micro$matrix=="soil"),]

#need to get actual microsome data and run michaelis menton kinetics
```

import experimental microsomal data
=================================
This data is observed from 0-90 minutes and has 3-4 replicates for each different micromolar concentration in a traditional Michaelis-Menton setup. Used to estimate slopes.

exposure experiment analysis
============================
Results presentation and discussion for parent analytes (atrazine, triadimenon, and fipronil) and their xenobiotic metabolites.

Toxicity of parents and metabolites discussion.

Microsomal analysis of soil and amphibian data for 0 (soil only), 2, 4, 12, 24, and 48 hours after exposure.

Database has factors for time, parent (mapped to analyte), analyte (can be either parent or metabolite), matrix (amphibian or soil), and tank (potentially a nuisance variable).

```{r eval=TRUE, echo=FALSE}
#using dplyr
micro.group <- group_by(micro, parent, analyte, matrix, time)
str(micro.group)
micro.group
micro.group.stats <- summarise(micro.group, 
            count = n(),
            ConcMean = mean(conc),
            ConcSD = sd(conc)
)

micro.group.stats
micro.group.stats.html <- print.xtable(xtable(micro.group.stats), type="html")
dput(micro.group.stats.html, file = paste(micro.tables,"micro.group.stats.html"
                                          ,sep=""))
#View(micro.group.stats)
```

The amphibian data set summary statistics. Atrazine peaks at 4 hours and then declines monotonically. DEA peaks at 12 hours and DIA at 24. Fipronil steadily declines from its first observation at 2 hours with fipronil sulfone not peaking until 24 hours. Triadimenon peaks at 24 hours with its metabolites tdla and tdlb peaking at 4 hours.

```{r eval=TRUE, echo=FALSE}
str(micro.group.stats)
micro.group.stats.amphib <- micro.group.stats[which(micro.group.stats$matrix=="amphib"),]
head(micro.group.stats.amphib)
#atrazine
micro.group.stats.amphib.atrazine <- micro.group.stats.amphib[which(micro.group.stats.amphib$parent=="atrazine"),]
micro.group.stats.amphib.atrazine
atrazine.analytes <- unique(micro.group.stats.amphib.atrazine$analyte)
for(analyte in atrazine.analytes){
  print(analyte)
}

#fipronil
micro.group.stats.amphib.fipronil <- micro.group.stats.amphib[which(micro.group.stats.amphib$parent=="fipronil"),]
micro.group.stats.amphib.fipronil
#triadimenon
micro.group.stats.amphib.triadimenon <- micro.group.stats.amphib[which(micro.group.stats.amphib$parent=="triadimenon"),]
micro.group.stats.amphib.triadimenon
```

```{r eval=TRUE, echo=FALSE}
#plot with zoo objects
```

The soil data set summary statistics. Atrazine in soil showing an upwards trend (need to test if significant) with dia and dea levels very low indicating little degradation in soil over the 48 hour period. Fipronil and triadimenon data also indicating little degradation in soil. 

```{r eval=TRUE, echo=FALSE}
micro.group.stats.soil <- micro.group.stats[which(micro.group.stats$matrix=="soil"),]
#atrazine
micro.group.stats.soil.atrazine <- micro.group.stats.soil[which(micro.group.stats.soil$parent=="atrazine"),]
micro.group.stats.soil.atrazine
#fipronil
micro.group.stats.soil.fipronil <- micro.group.stats.soil[which(micro.group.stats.soil$parent=="fipronil"),]
micro.group.stats.soil.fipronil
#triadimenon
micro.group.stats.soil.triadimenon <- micro.group.stats.soil[which(micro.group.stats.soil$parent=="triadimenon"),]
micro.group.stats.soil.triadimenon
```



```{r eval=TRUE, echo=TRUE}
pdf(paste(micro.graphics,"data_mean_scatterplot",".pdf", sep=""))
  par(mfrow=c(3,1))
  print(parents)
  for(parent in parents){
    i=0
    print(parent)
    temp.parent <- micro.group.stats.amphib[which(micro.group.stats.amphib$parent==parent),]
    print(temp.parent)
    parent.analytes <- unique(temp.parent$analyte)
    for(analyte in parent.analytes){
      print(parent)
      print(analyte) 
      analytetemp <- micro.group.stats.amphib[which(micro.group.stats.amphib$analyte==analyte),]
      xvalues <- as.numeric(as.character(analytetemp$time))
      points.y <- micro.amphib[which(micro.amphib$analyte==analyte),]$conc
      points.x <- as.numeric(as.character(micro.amphib[which(micro.amphib$analyte==analyte),]$time))
      #create empty plot if needed
      if(i==0){
        parenttemp <- micro.group.stats.amphib[which(micro.group.stats.amphib$parent==parent),]
        maxconc <- max(points.y) 
        plot(xvalues,analytetemp$ConcMean,type="l", xlim=c(0,48),
             ylim = c(0,maxconc), main=parent,col="black", xlab="Hours", ylab="Concentration")
        axis(1,at=xvalues)
        i=1
      }
      #add line for parent metabolite
      if(analyte %in% parents){
        lines(xvalues,analytetemp$ConcMean,type="l",col="red")
        points(points.x, points.y, col = "red")
      }else{
      #add line for daughter metabolite
        lines(xvalues,analytetemp$ConcMean,type="l",col="blue")
        points(points.x, points.y, col = "blue")
      }
    }
  }
dev.off()
```

la de da

#turn times series to zoo objects



#estimate lambdas using maximum likelihood for each time series
library(maxLik)

MLexp <- function(times, data){
  
  expLik <- function(param) {
    y <- data
    t <- times
    alpha1 <- param[1]
    lambda1 <- param[2]-1*(alpha1-lambda1*t) - y/(exp(alpha1-lambda1*t))
  }
  
  max.fit <- maxLik(expLik, start = c(1,1))
  crit <- qt(.975, length(data)-2)
  max.intercept.CI <- c(summary(max.fit)$est[1,1]-crit*summary(max.fit)$est[1,2],summary(max.fit)$est[1,1]+crit*summary(max.fit)$est[1,2])
  max.decayrate.CI <- c(-1*summary(max.fit)$est[2,1]-crit*summary(max.fit)$est[2,2],-1*summary(max.fit)$est[2,1]+crit*summary(max.fit)$est[2,2])
  max.halflife.CI <- c(log(2)/max.decayrate.CI[2],log(2)/max.decayrate.CI[1])
  
  paste("MLE for Initial Concentration =",max.fit$est[1],
        "MLE for decay rate =", max.fit$est[2],
        "MLE for half life =", -log(.5)/max.fit$est[2])
  
  temp.list <- list(max.fit$est[1],  max.fit$est[2], -log(.5)/max.fit$est[2], max.intercept.CI, max.decayrate.CI, max.halflife.CI, vcov(max.fit) )
  return(temp.list)
}

MLexp.intercept <- function(times, data) {
  temp.fit <- MLexp(times,data)
  temp.fit[[1]]
}

MLexp.decayrate <- function(times, data) {
  temp.fit <- MLexp(times,data)
  temp.fit[[2]]
}

MLexp.halflife <- function(times,data) {
  temp.fit <- MLexp(times,data)
  temp.fit[[3]]
}

MLexp.intercept.CI <- function(times,data){
  temp.fit <- MLexp(times,data)
  temp2 <- temp.fit[[4]]
  names(temp2) <- c("Lower95", "Upper95")
  temp2
}

MLexp.decayrate.CI <- function(times,data){
  temp.fit <- MLexp(times,data)
  temp2 <- temp.fit[[5]]
  names(temp2) <- c("Lower95", "Upper95")
  temp2
}

MLexp.halflife.CI <- function(times,data){
  temp.fit <- MLexp(times,data)
  temp2 <- temp.fit[[6]]
  names(temp2) <- c("Lower95", "Upper95")
  temp2
}

microsome experiment analysis
============================
data in csv format