microsome analyses
========================================================
Results presentation and discussion for parent analytes (atrazine, triadimenon, and fipronil) and their xenobiotic metabolites.

Toxicity of parents and metabolites discussion.

```{r eval=TRUE, echo=TRUE}
R.Version()$version.string
library(dplyr)
library(knitr)
library(grofit)
library(zoo)


#### read in and format data ####
if(Sys.info()[4]=="DC2626UTPURUCKE"){
  microdir <- "C:\\stp_drop\\Dropbox\\amphib_microsomes_glinskietal2015\\RCode\\"
}

if(Sys.info()[4]=="stp-air.local"){
  microdir <- "C:\\stp_drop\\Dropbox\\amphib_microsomes_glinskietal2015\\RCode\\"
}

micro <- read.table(paste(microdir,"Microsomes2.csv",sep=""), header = TRUE, sep = ",")
str(micro)
micro$time <- factor(micro$time)
micro$tank <- factor(micro$tank)
parents <- unique(micro$parent)
analytes <- unique(micro$analyte)
```

Microsomal analysis of soil and amphibian data for 0 (soil only), 2, 4, 12, 24, and 48 hours after exposure.

```{r eval=TRUE, echo=TRUE}
micro
parents
analytes
```

Database has factors for time, parent (mapped to analyte), analyte (can be either parent or metabolite), matrix (amphibian or soil), and tank (potentially a nuisance variable).

```{r eval=TRUE, echo=FALSE}
#using dplyr
micro.group <- group_by(micro, parent, analyte, matrix, time)
str(micro.group)
micro.group
micro.group.stats <- summarise(micro.group, 
            count = n(),
            ConcMean = mean(conc),
            ConcSD = sd(conc)
)

micro.group.stats



```

The amphibian data set summary statistics. Atrazine peaks at 4 hours and then declines monotonically. DEA peaks at 12 hours and DIA at 24. Fipronil steadily declines from its first observation at 2 hours with fipronil sulfone not peaking until 24 hours. Triadimenon peaks at 24 hours with its metabolites tdla and tdlb peaking at 4 hours.

```{r eval=TRUE, echo=FALSE}
str(micro.group.stats)
micro.group.stats.amphib <- micro.group.stats[which(micro.group.stats$matrix=="amphib"),]
head(micro.group.stats.amphib)
#atrazine
micro.group.stats.amphib.atrazine <- micro.group.stats.amphib[which(micro.group.stats.amphib$parent=="atrazine"),]
micro.group.stats.amphib.atrazine
atrazine.analytes <- unique(micro.group.stats.amphib.atrazine$analyte)
for(analyte in atrazine.analytes){
  print(analyte)
}

#fipronil
micro.group.stats.amphib.fipronil <- micro.group.stats.amphib[which(micro.group.stats.amphib$parent=="fipronil"),]
micro.group.stats.amphib.fipronil
#triadimenon
micro.group.stats.amphib.triadimenon <- micro.group.stats.amphib[which(micro.group.stats.amphib$parent=="triadimenon"),]
micro.group.stats.amphib.triadimenon
```

```{r eval=TRUE, echo=FALSE}
#plot with zoo objects
```

The soil data set summary statistics. Atrazine in soil showing an upwards trend (need to test if significant) with dia and dea levels very low indicating little degradation in soil over the 48 hour period. Fipronil and triadimenon data also indicating little degradation in soil. 

```{r eval=TRUE, echo=FALSE}
micro.group.stats.soil <- micro.group.stats[which(micro.group.stats$matrix=="soil"),]
#atrazine
micro.group.stats.soil.atrazine <- micro.group.stats.soil[which(micro.group.stats.soil$parent=="atrazine"),]
micro.group.stats.soil.atrazine
#fipronil
micro.group.stats.soil.fipronil <- micro.group.stats.soil[which(micro.group.stats.soil$parent=="fipronil"),]
micro.group.stats.soil.fipronil
#triadimenon
micro.group.stats.soil.triadimenon <- micro.group.stats.soil[which(micro.group.stats.soil$parent=="triadimenon"),]
micro.group.stats.soil.triadimenon
```



```{r eval=TRUE, echo=TRUE}
par(mfrow=c(3,1))
print(parents)
for(parent in parents){
  i=0
  print(parent)
  temp.parent <- micro.group.stats.amphib[which(micro.group.stats.amphib$parent==parent),]
  print(temp.parent)
  parent.analytes <- unique(temp.parent$analyte)
  for(analyte in parent.analytes){
    print(parent)
    print(analyte) 
    analytetemp <- micro.group.stats.amphib[which(micro.group.stats.amphib$analyte==analyte),]
    xvalues <- as.numeric(as.character(analytetemp$time))
    #create empty plot if needed
    if(i==0){
      parenttemp <- micro.group.stats.amphib[which(micro.group.stats.amphib$parent==parent),]
      maxconc <- max(parenttemp$ConcMean) 
      plot(xvalues,analytetemp$ConcMean,type="l", xlim=c(0,48),
           ylim = c(0,maxconc), main=parent,col="black")
      axis(1,at=xvalues)
      i=1
    }
    #add line for parent metabolite
    if(analyte %in% parents){
      lines(xvalues,analytetemp$ConcMean,type="l",col="red")
    }else{
    #add line for daughter metabolite
      lines(xvalues,analytetemp$ConcMean,type="l",col="blue")
    }
  }
}
```

la de da

#turn times series to zoo objects



#estimate lambdas using maximum likelihood for each time series
library(maxLik)

MLexp <- function(times, data){
  
  expLik <- function(param) {
    y <- data
    t <- times
    alpha1 <- param[1]
    lambda1 <- param[2]-1*(alpha1-lambda1*t) - y/(exp(alpha1-lambda1*t))
  }
  
  max.fit <- maxLik(expLik, start = c(1,1))
  crit <- qt(.975, length(data)-2)
  max.intercept.CI <- c(summary(max.fit)$est[1,1]-crit*summary(max.fit)$est[1,2],summary(max.fit)$est[1,1]+crit*summary(max.fit)$est[1,2])
  max.decayrate.CI <- c(-1*summary(max.fit)$est[2,1]-crit*summary(max.fit)$est[2,2],-1*summary(max.fit)$est[2,1]+crit*summary(max.fit)$est[2,2])
  max.halflife.CI <- c(log(2)/max.decayrate.CI[2],log(2)/max.decayrate.CI[1])
  
  paste("MLE for Initial Concentration =",max.fit$est[1],
        "MLE for decay rate =", max.fit$est[2],
        "MLE for half life =", -log(.5)/max.fit$est[2])
  
  temp.list <- list(max.fit$est[1],  max.fit$est[2], -log(.5)/max.fit$est[2], max.intercept.CI, max.decayrate.CI, max.halflife.CI, vcov(max.fit) )
  return(temp.list)
}

MLexp.intercept <- function(times, data) {
  temp.fit <- MLexp(times,data)
  temp.fit[[1]]
}

MLexp.decayrate <- function(times, data) {
  temp.fit <- MLexp(times,data)
  temp.fit[[2]]
}

MLexp.halflife <- function(times,data) {
  temp.fit <- MLexp(times,data)
  temp.fit[[3]]
}

MLexp.intercept.CI <- function(times,data){
  temp.fit <- MLexp(times,data)
  temp2 <- temp.fit[[4]]
  names(temp2) <- c("Lower95", "Upper95")
  temp2
}

MLexp.decayrate.CI <- function(times,data){
  temp.fit <- MLexp(times,data)
  temp2 <- temp.fit[[5]]
  names(temp2) <- c("Lower95", "Upper95")
  temp2
}

MLexp.halflife.CI <- function(times,data){
  temp.fit <- MLexp(times,data)
  temp2 <- temp.fit[[6]]
  names(temp2) <- c("Lower95", "Upper95")
  temp2
}

